---
title: "Assessment"
author: "Mary"
date: "2024-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Adding libraries

```{r, include=FALSE}
library(tidyverse)
library(janitor)
library(gt)
library(here)
library(kableExtra)
```

## Introduction

It this report I will be focusing on the number of prescribed doses of medication for ADHD per person. I will compare data before the first COVID-19 lockdown in the UK (march 2020) and the most recent data (july 2024). I will try to find any trends and differences between these two datasets. The aim of this report is to answer the question of: Is there an increased prescription of ADHD drugs pre and post COVID-19? The reason I am comparing these two datapoints is that there are rumours that quarantine lowered people's attention span as everyone was learning and working from home, looking at their computer the whole day. I would like to find out if this is true and if there was an increase in prescription of ADHD drugs to tackle the problem of low attention span.

# Let's load our data and filter for 5 drugs used to treat ADHD

```{r, include=FALSE}
july_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/1cb425fc-640c-4b37-9013-f8e97f274085/download/pitc202407.csv") %>% 
  filter(str_detect(BNFItemDescription, "METHYLPHENIDATE|LISDEXAMFETAMINE|DEXAMFETAMINE|ATOMOXETINE|GUANFACINE"))

march_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9581bd8d-5568-4462-93a6-6d7bfbbe7cbc/download/pitc202003.csv") %>% 
  filter(str_detect(BNFItemDescription, "METHYLPHENIDATE|LISDEXAMFETAMINE|DEXAMFETAMINE|ATOMOXETINE|GUANFACINE"))
```

## Including Plots

You can also embed plots, for example:

```{r pressure, include=FALSE}
pressure %>% 
  ggplot(aes(x = temperature, y = pressure)) +
  geom_point() +
  theme_bw()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
